Optical characterization of gaps in directly bonded Si compound optics using infrared spectroscopy

### 2015-03-23 AO_bondedSi.tex
incorporate DTJ edits up to page 17

In the limit $\lim_{d \to 0} T_g \rightarrow T_{DSP} $ and the gap approaches 100\% transmission.  Figure \ref{figAbsoluteTrans} shows a plot of Equation \ref{eqFP} for gap sizes, $d$, of 0, 20, 35, 60, 350, and 3500 nm.  The bottom two panels show the ``gap'' transmission, the value of the throughput with the losses from the outer Si cavities divided out.  For the very largest gap, large fringes appear as the cavity resonance of the air gap modulates the transmission.  Below a $\lambda/2$ gap size the effect of the gap manifests itself as a wavelength-dependent diminution of the transmission.  For the smallest $35-60$ nm gaps, there is a $2-6\%$ decrease in gap transmission from $2400\;$nm down to $1250\;$nm.  It is the magnitude and wavelength dependence of this transmission spectrum that provides information about the axial extent of the gap.

@@

We evaluated our metrology method by embedding gaps of known sizes in the interface and measuring the effects of these gaps on IR transmission.  The substrate thicknesses ranged from 0.5 to 3.3 mm.  We denote substrates with thicknesses larger than 1 mm as ``pucks'' and substrates with thickness less than 1.0 mm thick as ``wafers''.  We distinguish the two classes of substrate thickness based on the anticipation that thin substrates will conform more easily to their bonding partner substrates.  There is also evidence that bond front propagation is slower in thick pucks \cite{2007ApOpt..46.6793H} than in wafers.  Our ultimate goal is to bond optics with thicknesses as large as 30 mm.

The three important characteristics of gaps are their axial extent, their lateral areal extent, and their areal fill factor.  To fabricate our test samples, we used photoresist lithography to pattern Si substrates, and we bored holes with inductively coupled plasma etching.  We used two different gases with low and high etch rates, CHF$_3$ on Si exhibited about 0.3 nm/second etch rate, SF$_6$ exhibited about 1 micron/minute.  Tables \ref{tbl_experiments} and \ref{tbl_meshPatterns} give information about the hole patterns and depths, as measured with Veeco NT9100 Optical Profiler for the small depth, and Dektak stylus profilometry for the large depth.

The fill factor is defined as the pattern area covered by gaps relative to the total pattern area.  In a second round of experiments, we achieved gap sizes of $14-95\;$nm on four substrates with three meshes with coarse, medium, and fine boxes, each with 50\% fill factors.  No attempt was made to verify the delivered fill factor, but since high precision lithographic processes were employed, we assign no uncertainty to the difference between the delivered fill factor and the designed fill factor.  For measurements taken with the Veeco NT9100 Optical Profiler, uncertainties were constructed by inspecting the histogram of topology, as shown in Figure \ref{figVS20pattern}.  The large-scale distortions were removed from the topology by masking and flattening to ensure that the uncertainties accurately reflect the distribution of measured heights.  The stylus profilometry measured values were simply assigned an uncertainty of 5\%, which was consistent with our intuition and experience.  Table \ref{tbl_experiments} lists part numbers and descriptions for our bonded substrates.

@@

We cleaned the surfaces before bonding to minimize the interfacial particle density.  The surface roughness was typically about 2 nm, as measured with a Veeco NT9100 Optical Profiler.  We measured the large scale surface flatness of the pucks.  The Fisba2 interferometer used for these measurements has a 50 mm diameter beam at $\lambda=632.8\;$nm.  We found a typical peak to valley surface flatness of 2 waves over the central 50 mm diameter.  We prepared the substrates with standard cleaning procedures of solvents in a megasonic.  We then applied MHz frequency oxygen plasma ashing.  We soaked the wafers in DI water then dried with N$_2$.  We pressed the Si substrates together from the center to the outside.

\section{Measurement technique and correlated measurement errors}
\label{sec_aboutErrors}
We took transmission spectra of bonded Si wafers and pucks, and double side polished Si reference wafers using an Agilent Cary 5000 UV-Vis-NIR spectrophotometer.  Table \ref{tabCary5000pars} lists the typical measurement settings and tool performance.  The tool has a double monochromator, which reduces scattered light.  The vendor reported linearity exceeds 40 dB.

The Cary 5000 has four operation modes: single front, single back, double, and double reverse.  We experimented with single front and double modes.  Double mode takes a reference spectrum simultaneously with the target spectrum.  In both single and double mode, a baseline scan is taken with no sample in the measurement holder.  We automatically divide sample spectra by the baseline spectrum.  In single beam mode, the baseline measurements must be repeated about every 10 minutes for post-process division.  In double beam mode the front beam is automatically divided by the rear beam, which corrects out lamp luminosity drift, for example.

We estimated the uncertainty and measurement repeatability of single and double modes in the following way: we computed the RMS error and measurement repeatability in a single measurement by computing the standard deviation of a spectrum taken immediately after a baseline scan with no sample present.  We found that the mean value in single mode drifted by about 0.2\% over several measurements.  The standard deviation of the featureless spectrum was typically 0.02\% from point to point, with some portions of the spectrum demonstrating increased variance, perhaps attributable to atmospheric absorption.  In double beam mode, the spectrograph drift was much lower.  The simultaneous reference in double beam mode divides out the 0.2\% drift in the mean transmission level attributable to lamp drift that we saw in single beam measurements.  In double beam mode, the drift in mean transmission level is 0.02\%.  These values are accurate for measurements over a half hour timescale or less.  Table \ref{tabCary5000pars} summarizes the amplitudes of the drifts and individual measurement errors in both single and double beam modes.\emph{Correlated errors dominate the measurement uncertainty of our transmission spectra.}  The measurement uncertainty correlations are apparent in in Figure \ref{figSiAbsorbfig}.  In some parts of the spectrum, the mean drift is about ten times larger than the standard deviation of the spectrum.  Double beam mode has about ten times lower lamp drift but suffers from other instrumental artifacts, like disjointed spectra at the wavelengths of known grating changeover points.

The spectral sampling ranged from 2.0 to 10.0 nm.  We used a spectral resolution of 5.0 nm.  We expect smoothly varying spectral features for all conceivable interface gaps in Si bonded wafers.  Specifically, Fabry-P\`erot fringes attributable to interfacial gaps in bonded Si will be well-sampled with 5 nm spectral resolution, so long as the interface gap is less than $\sim$100 $\mu$m.  The axial extents of such large gaps would be easily measurable by other means, like IR imaging.

@@

\subsection{IR bubble detection and limitations}

We looked for large gaps, detectable as ``IR bubbles'' \cite{1992JEMat..21..669M} with IR imaging.  The detector was an IR Vista $\alpha$NIR infrared focal plane array, with 312 $\times$ 252 pixels, with 30 $\mu$m square pixels.  We used a 50 mm F/2 lens, with a fiber optic white light illumination source.  The combination of the absorption of Si, the lamp spectrum and the detector responsivity led to an effective bandpass from approximately $1.15-1.6\;\mu$m.  The field of view was smaller than the 100 mm diameter wafers, so we dithered the sample.  Figure \ref{figVS2021_IR_image} shows an IR image of bonded sample VS20-21.  For this image, we coarsely flat-fielded the detector by observing a homogeneously illuminated white screen. Despite the flat-fielding, detector non-uniformities and vignetting are perceptible in our image as vertical stripes.  Mosaic-stitching errors are also perceptible.  It was easy to detect IR bubbles despite detector artifacts since the bubbles translate with the sample when we dither its position.

We also used IR imaging to inspect VG09-12 and VG15-13.  We detected IR bubbles in all 3 of these bonded Si samples.  The observable bubble areal density varied from about 4 per 100 mm diameter bonded wafer pair to over 20 per bonded wafer pair.  The individual IR bubble areas were as small as a few square millimeters up to $400\;\mathrm{mm}^2$.  The IR bubbles had varied morphologies.  The structure within the IR bubbles was sometimes indicative of Newton's rings, exposing the presence of gaps larger than about $\lambda/2$ up to $\sim 15 \lambda$.  As expected, the unbonded DSP wafers show no such IR bubbles.  No attempt at quantitative measurement was made on the IR images due to evidence of a non-linear detector response.  The bulk morphologies were clear regardless of the detector response.  When IR imaging was available, it was leveraged to target IR specectroscopy in locations demonstrating presence or absense of IR bubbles, as desired.  In other cases when IR imaging was not available or ignored, spectra were taken at random positions on the wafer.

@@

The gaps in our test pairs should be a combination of the deliberately induced gaps and any inadvertent additional spaces caused by imperfections in the bonds.  We infer the axial extent of gaps in our samples with etched holes from their measured transmission spectra.  We use a Bayesian inference technique.  First we define a mixture model in which the observed normalized gap spectrum $T_{obs}$ is composed of a sum of a spectrum through a gap of axial extent $d$, and a spectrum through a perfect bond:

@@

where $f$ is the areal fill factor of the region exhibiting a finite gap size, and $1-f$ is therefore the areal fill factor of the perfectly bonded area.  $T_g$ is defined in equation \ref{eqn:Tetalon}.  The only free parameters in Equation \ref{eqnMix} are $d$ and $f$.  Both of these parameters are fixed in our direct bonded Si wafers with synthetic gaps, and are listed in Table \ref{tbl_meshPatterns}.  These synthetic imbedded gaps therefore offer an excellent test of our ability to recover the gap axial extent $d$ and areal fill factor $f$ (Section \ref{secKnownGaps}).

>>>>

In the limit $\lim_{d \to 0} T_g \rightarrow T_{DSP} $ and the gap approaches 100\% transmission.  Figure \ref{figAbsoluteTrans} shows a plot of Equation \ref{eqFP} for gap sizes, $d$, of 0, 20, 35, 60, 350, and 3500 nm.  The bottom two panels show the ``gap'' transmission based on Equation \ref{eqFP}, the value of the throughput with the losses from the outer Si cavities divided out.  For the very largest gap, large fringes appear as the cavity resonance of the air gap modulates the transmission.  Below a $\lambda/2$ gap size, the gap manifests itself as a wavelength-dependent diminution of the transmission.  For the smallest ($35-60$ nm) gaps, there is a $2-6\%$ decrease in gap transmission from $2400\;$nm down to $1250\;$nm.  It is the magnitude and wavelength dependence of this transmission spectrum that provides information about the axial extent of the gap.

@@

We evaluated our metrology method by embedding gaps of known sizes in the interface between two bonded Si substrates and measuring the effects of these gaps on IR transmission.  The substrate thicknesses ranged from 0.5 to 3.3 mm.  We denote substrates with thicknesses larger than 1 mm as ``pucks'' and substrates with thickness less than 1.0 mm thick as ``wafers''.  We distinguish the two classes of substrate thickness based on the anticipation that thin substrates will conform more easily to their bonding partner substrates.  There is also evidence that bond front propagation is slower in thick pucks \cite{2007ApOpt..46.6793H} than in wafers.  Our ultimate goal is to bond optics with thicknesses as large as 30 mm.

The three important characteristics of gaps are their axial extent, their lateral areal extent, and the areal fill factor of ensembles of such gaps.  To fabricate our test samples, we used photoresist lithography to pattern Si substrates, and we bored holes with inductively coupled plasma etching.  We used two different gases with low and high etch rates, CHF$_3$ on Si exhibited about 0.3 nm/second etch rate, SF$_6$ exhibited about 1 $\mu$m/minute to produce gaps with small and large depths.  Tables \ref{tbl_experiments} and \ref{tbl_meshPatterns} give information about the hole patterns and depths, as measured with Veeco NT9100 Optical Profiler for the small depth, and Dektak stylus profilometry for the large depth.

The fill factor is the pattern area covered by gaps relative to the total pattern area.  We achieved gap sizes of $14-95\;$nm on four substrates with three meshes with coarse, medium, and fine boxes, each with 50\% fill factors.  We did not verify the delivered fill factor, but since high precision lithography is very reliable, we assign no uncertainty to the difference between the delivered fill factor and the designed fill factor.  For measurements taken with the Veeco NT9100 Optical Profiler, uncertainties were constructed by inspecting the histogram of topology, as shown in Figure \ref{figVS20pattern}.  The large-scale distortions were removed from the topology by masking and flattening to ensure that the uncertainties accurately reflect the distribution of measured heights.  The stylus profilometry measured values were assigned an uncertainty of 5\%, our experience with other etched parts.  Table \ref{tbl_experiments} lists part numbers and descriptions for our bonded substrates.

@@

We cleaned the surfaces before bonding to minimize the interfacial particle density.  The surface roughness was typically about 2 nm, as measured with a Veeco NT9100 Optical Profiler.  We also measured the large scale surface flatness of the pucks.  The Fisba2 interferometer used for these measurements has a 50 mm diameter beam at $\lambda=632.8\;$nm.  We found a typical peak to valley surface flatness of 2 waves over the central 50 mm diameter.  We prepared the substrates with standard cleaning procedures of solvents in a megasonic.  We then applied MHz frequency oxygen plasma ashing.  We soaked the wafers in DI water then dried with N$_2$.  We pressed the Si substrates together from the center to the outside.

\subsection{IR bubble imaging and limitations}

We first looked for large gaps in our bonded substrate pairs detectable as ``IR bubbles'' \cite{1992JEMat..21..669M}, with IR imaging.  The detector was an IR Vista $\alpha$NIR infrared focal plane array, with 312 $\times$ 252 pixels, with 30 $\mu$m square pixels.  We used a 50 mm F/2 lens, with a fiber optic white light illumination source.  The combination of the absorption of Si, the lamp spectrum and the detector responsivity led to an effective bandpass from approximately $1.15-1.6\;\mu$m.  The field of view was smaller than the 100 mm diameter wafers, so we dithered the sample.  Figure \ref{figVS2021_IR_image} shows an IR image of bonded sample VS20-21.  For this image, we coarsely flat-fielded the detector by observing a homogeneously illuminated white screen. Despite the flat-fielding, detector non-uniformities and vignetting are perceptible in our image as vertical stripes.  Mosaic-stitching errors are also perceptible.  It was easy to detect IR bubbles despite detector artifacts since the bubbles translate with the sample when we dither its position.

We also used IR imaging to inspect VG09-12 and VG15-13.  We detected IR bubbles in all 3 of these bonded Si samples.  The observable bubble areal density varied from about 4 per 100 mm diameter bonded wafer pair to over 20 per bonded wafer pair.  The individual IR bubble areas were as small as a few square millimeters up to $400\;\mathrm{mm}^2$.  The IR bubbles had various morphologies.  The structure within some of the images of IR bubbles showed Newton's rings, exposing the presence of gaps larger than about $\lambda/2$ up to $\sim 15 \lambda$.  As expected, the unbonded DSP wafers show no such IR bubbles.  No attempt at quantitative measurement was made on the IR images due to evidence of a non-linear detector response.  The bulk morphologies were clear regardless of the detector response.  When IR imaging was available, it was leveraged to target IR specectroscopy in locations demonstrating presence or absense of IR bubbles, as desired.  In other cases when IR imaging was not available or ignored, spectra were taken at random positions on the wafer.

\section{Measurement technique and correlated measurement errors}
\label{sec_aboutErrors}

We used the test pieces with known gaps to validate our measurement and analysis techniques.  We took transmission spectra of the bonded Si wafers and pucks, and the double side polished Si reference wafers using an Agilent Cary 5000 UV-Vis-NIR spectrophotometer.  Table \ref{tabCary5000pars} lists the typical measurement settings and tool performance.  The tool has a double monochromator, which reduces scattered light.  The vendor reported linearity exceeds 40 dB.

The Cary 5000 has four operation modes: single front, single back, double, and double reverse.  Double mode takes a reference spectrum simultaneously with the target spectrum.  We experimented with single front and double modes.  In principle double beam mode is even more precise than single beam mode.  In practice we found that souble beam mode showed evidence for lamp-jump artifacts attributable to polarization differences between the two arms.  So in this article we focus our results to those obtained in single beam mode.  In both single and double mode, the spectrometer takes a baseline scan with no sample in the measurement holder.  We automatically divide sample spectra by this baseline spectrum.  In single beam mode, we must repeat the baseline measurements about every 10-30 minutes for post-process division.

We estimated the uncertainty and measurement repeatability experimentally.  We computed the RMS error and measurement repeatability in a single measurement by computing the standard deviation of a spectrum taken immediately after a baseline scan with no sample present.  We found that the mean value in single mode drifted by about 0.2\% over several measurements.  The standard deviation of the featureless spectrum was typically 0.02\% from point to point, with some portions of the spectrum demonstrating increased variance.  Table \ref{tabCary5000pars} summarizes the amplitudes of the drifts and individual measurement errors.  \emph{Correlated errors dominate the measurement uncertainty of our transmission spectra.}  The measurement uncertainty correlations are apparent in Figure \ref{figSiAbsorbfig}.  In some parts of the spectrum, the mean drift is about ten times larger than the local standard deviation of the spectrum.

The spectral sampling ranged from 2.0 to 10.0 nm.  We set the spectrometer slit width to produce a spectral resolution of 5.0 nm.  We expect smoothly varying spectral features for all conceivable interface gaps in Si bonded wafers.  Specifically, Fabry-P\`erot fringes attributable to interfacial gaps in bonded Si will be well-sampled with 5 nm spectral resolution, so long as the interface gap is less than $\sim$100 $\mu$m.  This maximum gap limitation brings our technique to the point where it overlaps in capability with other techniques since the axial extents of such large gaps would be easily measurable by other means, like IR imaging and fringe-counting.

We also verified the linearity of the detector response my measuring the transmission through stacks of filters.  We constained the linearity to at least $>24$ dB.  The vendor reports linearity exceeding 40 dB, which is consistent with our measurement.

@@

@@

The gaps in our test pairs should be a combination of the deliberately induced gaps and any inadvertent additional spaces caused by imperfections in the bonds.  We infer the axial extent of gaps in our samples with etched holes from their measured transmission spectra.  We use a Bayesian inference technique in our analysis.  First we define a mixture model in which the observed normalized gap spectrum $T_{obs}$ is composed of a sum of a spectrum through a gap of axial extent $d$, and a spectrum through a perfect bond:

@@

where $f$ is the areal fill factor of the region exhibiting a finite gap size, and $1-f$ is therefore the areal fill factor of the perfectly bonded area.  $T_g$ is defined in Equation \ref{eqFP}.  The only free parameters in Equation \ref{eqnMix} are $d$ and $f$.  Both of these parameters are fixed in our direct bonded Si wafers with synthetic gaps, and are listed in Table \ref{tbl_meshPatterns}.  These synthetic imbedded gaps therefore offer an excellent test of our ability to recover the gap axial extent $d$ and areal fill factor $f$ (Section \ref{secKnownGaps}).
